---
title: "README"
author: "Gregory J. Botwin"
date: "2/21/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
library(knitr)
library(tidyverse)
```

## Project Goals
I want to study phenotypes that are genetically similar to IBD using the UK Biobank Phentoype Data produced by the Neale Lab http://www.nealelab.is/blog/2017/7/19/rapid-gwas-of-thousands-of-phenotypes-for-337000-samples-in-the-uk-biobank 

## Tasks
### Get list of all IBD associated SNPs
On 2/21/18 I queried the NHGRI-EBI Catalog of published genome-wide association studies for:  
- Inflammatory Bowel Disease  
- p value <= **5x10^-8**
File saved in "data/gwas-association-downloaded_2018-02-21-inflammatory bowel disease.tsv"

A description of the terms can be found http://www.ebi.ac.uk/gwas/docs/methods 

```{r}
ibd_gwas_catalog <- read_tsv("data/gwas-association-downloaded_2018-02-21-inflammatory bowel disease.tsv") 
colnames(ibd_gwas_catalog) <- make.names(colnames(ibd_gwas_catalog))
kable(ibd_gwas_catalog %>%
  group_by(DISEASE.TRAIT) %>%
  summarise(n_snps = length(unique(STRONGEST.SNP.RISK.ALLELE))) %>%
  arrange(desc(n_snps)), caption = "Number of Unique SNPs with P <= 5x10^-8 per IBD Trait")
```

## Notes on the GWAS
### Sample Population 
- **487,409** individuals with phased and imputed genotype data 
- filter for British genetic ancestry
- filter closely related individuals (or at least one of a related pair of individuals) 
- filter individuals with sex chromosome aneuploidies
- filter individuals who had withdrawn consent from the UK Biobank study
- left with **337,199** genetically unqiue individuals with British ancestry

### Number of Variants
- Affymetrix UK BiLEVE Axiom array on an initial 50,000 
- Affymetrix UK Biobank Axiom array on remaining 450,000
- **820,967** SNP and indel markers on UK Biobank Axiom Array
- **92 million** imputed autosomal SNPs were available for analysis.
- Someone screwed up, half of these SNPs were undergoing re-imputation by the UK Biobank team due to mapping issues, leaving the ~40 autosomal million SNPs imputed from the Haplotype Reference Consortium as eligible for analysis.
- restricted to SNPs with minor allele frequency (MAF) > 0.1%
- restricted to SNPs with HWE p-value > 1e-10 
- leaving **10,894,596** million SNPs for analysis

### Association Tests (Neale Lab)
- least-squares linear model predicting the phenotype with an additive genotype coding (0, 1, or 2 copies of the minor allele) for all phenotypes
- sex and the first 10 principal components provided as covariates

### Post-Hoc QC
- Post-Hoc analysis by Neale lab found statistical inflation, elevated lambda GC, in case control analysis with small case size and large control size. 
- this inflation was dominant among the lowest frequency SNPs.
- Neale lab recommends to determine the allele frequency where we would expect at least 25 minor alleles in the smaller group 


















and to remove SNPs below that threshold

## Determine Genetic Similarity 
"For two diseases, A and B are associated with m mutations and n mutations, respectively. The chance of one mutation related with disease A in all mutations existed in the database is m/N, where N is the total number of variants in the UKBB. Thus, the expected number of mutations shared by both A and B is m × n/N..." DOI: 10.1002/humu.23358 
